---
title: "Figure - Compare Warming Distirbutions"
author: "Joe Brown"
date: "2025-06-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set-up 

```{r}
source("set-up/libraries.R")
```


We want to read in the metric result data for each experiment, add a column for the experiment code, and then cbind rows into a single dataframe.

```{r}
# Load files
dir_path <- "output/warming_results"
file_list <- list.files(path = dir_path, pattern = "\\.csv$", full.names = TRUE)

# Define mapping from exp to config
exp_map <- c("exp1" = "concentration",
             "exp2" = "concentration-emission",
             "exp3" = "emission")

# Read each file and assign config + term
metrics_list <- lapply(file_list, function(file) {
  df <- read.csv(file)
  
  # Remove 'X' column if it exists
  if ("X" %in% names(df)) {
    df <- df[ , names(df) != "X"]
  }
  
  file_name <- basename(file)
  
  # Extract experiment ID (e.g., exp1)
  exp_id <- sub("_.*", "", file_name)
  config <- exp_map[exp_id]
  
  # Determine term based on suffix in filename
  term <- if (grepl("mid-term", file_name)) {
    "mid-term"
  } else if (grepl("near-term", file_name)) {
    "near-term"
  } else {
    "long-term"  # No term in name → must be long-term
  }
  
  df$config <- config
  df$term <- term
  return(df)
})

# Combine all into one dataframe
combined_metrics <- do.call(rbind, metrics_list)
rownames(combined_metrics) <- NULL
```

# Start plotting

Now that the data is loaded and the formatting is corrected (which you will want to fix in the write csv code), we can move on to plotting the overlayed weighted warming distirbutions. 

```{r}
lt_warming_plot <- 
  ggplot() +
  stat_density(
    data = subset(combined_metrics, term == "long-term"), 
    aes(x = metric_result, weight = mc_weight, color = config), 
    geom = "line", 
    linewidth = 1,
    adjust = 1
  ) +
  labs(
    x = "2081-2100 Temperature Anomaly (C)",
    y = "Density (weighted)") +
  theme_light() +
  theme(axis.title.y = element_blank()) +
  facet_wrap(~scenario, scales = "free_x")

lt_warming_plot
```
### mid-term plot

```{r}
mt_warming_plot <- 
  ggplot() +
  stat_density(
    data = subset(combined_metrics, term == "mid-term"), 
    aes(x = metric_result, weight = mc_weight, color = config), 
    geom = "line", 
    linewidth = 1,
    adjust = 1
  ) +
  labs(
    x = "2041-2060 Temperature Anomaly (C)",
    y = "Density (weighted)"
  ) +
  theme_light() +
  theme(axis.title.y = element_blank()) +
  facet_wrap(~scenario, scales = "free_x")

mt_warming_plot
```
### near-term plot

```{r}
nt_warming_plot <- 
  ggplot() +
  stat_density(
    data = subset(combined_metrics, term == "near-term"), 
    aes(x = metric_result, weight = mc_weight, color = config), 
    geom = "line", 
    linewidth = 1,
    adjust = 1
  ) +
  labs(
    title = "Likelihood-weighted Warming Distribution",
    x = "2021-2040 Temperature Anomaly (C)",
    y = "Density (weighted)"
  ) +
  theme_light() +
  facet_wrap(~scenario, scales = "free_x")

nt_warming_plot
```
## Combine plots 

```{r}
library(ggpubr)

combined_pdf <- ggarrange(nt_warming_plot, mt_warming_plot, lt_warming_plot, 
                          ncol = 3, nrow = 1, 
                          common.legend = T, legend = "bottom")

combined_pdf

ggsave("combined_plot.png", combined_pdf, width = 15, height = 8)
```


# Violin plots

This is just a preliminary figure 

```{r}
ggplot(data = combined_metrics, 
       aes(x = config, y = metric_result, weight = mc_weight)) +
  geom_violin(scale = "area", fill = NA, color = "black") +
  geom_point(aes(size = mc_weight, alpha = mc_weight, color = config), position = position_jitter(width = 0.2), shape = 16) +
  scale_size_continuous(range = c(0.5, 3)) +
  scale_alpha_continuous(range = c(0.2, 0.8)) +
  labs(
    title = "End-of-century GMST projections",
    x = "Experiment",
    y = "GMST in 2100 (°C)"
  ) +
  facet_wrap(~scenario + term, scales = "free_y") +
  guides(size = "none", alpha = "none")+
  theme_light() +
  theme(axis.text.x = element_blank()) 

ggsave("temp_violin_comparison.png", height = 10, width = 15)
```

This figure illustrates how different treatments of historical atmospheric CO2 in simple climate models fundamentally alter the shape and spread of future warming projections. While most reduced-complexity climate models adopt a concentration-driven approach in the historical period, a strategy that ensures close alignment with observations and seems intuitively sound, this experiment reveals that doing so significantly narrows the projected uncertainty range. Fully emissions-driven simulations, by contrast, allow historical carbon cycle uncertainty to propagate forward, resulting in broader and more dispersed warming distributions. Even the hybrid setup, where CO2 concentrations are prescribed only during the historical period, dampens ensemble spread in a way that persists into the future. These differences arise not from parametric variation, but from the structural assumptions embedded in how we treat historical carbon dynamics. This finding underscores that concentration-driven setups, while practical and observationally anchored, come with the cost of suppressing uncertainty in the Earth system's response to emissions. Emissions-driven historical runs, despite appearing redundant in the presence of observational CO2 data, offer a critical view into the range of plausible carbon cycle behaviors that can seed future divergence. This experiment highlights that uncertainty in future warming is not just driven by emissions or physical parameters, but also by how we represent and constrain the past, a methodological choice that deserves more scrutiny in ensemble design and uncertainty communication.

Bullet point takeaways:

 - What the figure shows:
 
 Prescribing CO2 in both history and future yields narrow, peaked warming projections. Prescribing CO2 only in history (green) produces intermediate spread. Fully emissions-driven runs show broader, flatter distributions, exposing the influence of carbon cycle feedback uncertainty.

 - What might be novel here:
 
 Demonstrates, within the same model and ensemble, how structural choices in forcing setup alone shape projection uncertainty. Reveals that historical constraint methods (not just parameter values or emissions scenarios) meaningfully change outcomes. Highlights the downstream consequences of ensemble construction strategy — something rarely tested explicitly in RCMs.

 - Why this matters despite having CO2 observations:
 
 Using observed CO2 in history makes sense practically, but it removes all historical carbon cycle flexibility. Emissions-driven runs let us explore how uncertainty in land/ocean sinks and feedbacks could influence initial conditions and propagate forward. Even if we can match CO2 with concentrations, it's scientifically valuable to ask: what range of behaviors could have produced that curve?
 
 If we know what range of behaviors produced that curve, it is likely that those same behaviors will follow earth system physics into the future?

 - How this helps us understand SCM uncertainty:
 
 Shows that future projection spread is influenced not just by emissions or parameter choices, but by how we structure the model historically. Challenges the assumption that SCMs always underpredict uncertainty because of missing processes — sometimes it's from over-constraining history. Suggests that using a blend of concentration- and emissions-driven ensembles might better capture the full range of plausible Earth system responses.

## Computing summaries of end of century warming

Here I want to calculat the median and CI of the end of century warming. So I want to use the metric data, group it by experiment and scneario, and compute the weighted quntiles. 

First start this with the metric data then we can break into the time series data. 

```{r}
metric_summary <- 
  combined_metrics %>% 
  group_by(config, scenario, term) %>% 
  dplyr::summarize(
    ci_05 = weighted.quantile(metric_result, w = mc_weight, probs = c(0.05)), 
    ci_33 = weighted.quantile(metric_result, w = mc_weight, probs = c(0.33)), 
    ci_50 = weighted.quantile(metric_result, w = mc_weight, probs = c(0.50)), 
    ci_66 = weighted.quantile(metric_result, w = mc_weight, probs = c(0.66)),
    ci_95 = weighted.quantile(metric_result, w = mc_weight, probs = c(0.95)), 
  .groups = "drop")

```

```{r}
ggplot() +
  # Jittered points with weighted transparency and size
  geom_jitter(
    data = combined_metrics,
    aes(x = config, y = metric_result, size = mc_weight, alpha = mc_weight, color = config),
    width = 0.2,
    shape = 16
  ) +
  # Error bars from 5th to 95th
  geom_errorbar(
    data = metric_summary,
    aes(x = config, ymin = ci_05, ymax = ci_95),
    width = 0.1,
    linewidth = 0.5
  ) +
  # Box (33rd to 66th)
  geom_rect(
    data = metric_summary,
    aes(
      xmin = as.numeric(factor(config)) - 0.25,
      xmax = as.numeric(factor(config)) + 0.25,
      ymin = ci_33,
      ymax = ci_66,
      fill = config, 
      colour = config
    ),
    alpha = 0.6,
  ) +
  # Median line
  geom_segment(
    data = metric_summary,
    aes(
      x = as.numeric(factor(config)) - 0.25,
      xend = as.numeric(factor(config)) + 0.25,
      y = ci_50,
      yend = ci_50
    ),
    linewidth = 1
  ) +
  facet_wrap(~scenario + term, scales = "free_y") +
  scale_alpha_continuous(range = c(0.2, 0.8)) +
  scale_size_continuous(range = c(0.5, 2.5)) +
  labs(
    title = "Weighted GMST Distributions in 2100 by Experiment",
    x = "Experiment",
    y = "GMST in 2100 (°C)"
  ) +
  guides(size = "none", alpha = "none") +  
  theme_light() +
  theme(
    axis.text.x = element_blank(),         # remove x-axis labels
    axis.ticks.x = element_blank(),        # remove x-axis tick marks
    legend.position = "right"              # keep color legend visible
  )

ggsave("temp_box_comparison.png", height = 10, width = 15)
```

## Visualizing parameter sensitivtiy or something

This plot not very informative, thought it was going to tbe more interesting than it ended up being. 

```{r}
source("set-up/load_params.R")

parameters$run_number <- 1:nrow(parameters)

parameter_sensitivity_df <- dplyr::left_join(combined_metrics, parameters, by = "run_number")

param_sensitivity_long <- parameter_sensitivity_df %>%
  pivot_longer(cols = c(ECS, BETA, Q10_RH, NPP_FLUX0, AERO_SCALE, DIFFUSIVITY),  
               names_to = "parameter_name",
               values_to = "parameter_value")
```

```{r}
ggplot(param_sensitivity_long, aes(x = parameter_value, y = metric_result, color = config)) +
  geom_smooth(aes(weight = mc_weight), se = FALSE, method = "loess", linewidth = 1) +
  facet_wrap(~ scenario + parameter_name, scales = "free_x") +
  labs(
    title = "Parameter Sensitivity of GMST by Configuration",
    x = "Parameter Value",
    y = "GMST in 2100 (°C)",
    color = "Configuration"
  ) +
  theme_light() +
  theme(legend.position = "bottom")

ggsave("example_plot.png", height = 20, width = 11)
```


# Plotting comparisons of metrics across SSP scenarios and term lengths

First need to pivot the data that has my metric summary data
```{r}
wide_metric_summary <- metric_summary %>% 
  mutate(iqr = ci_05 - ci_95) %>% 
  select(config, scenario, term, ci_50, iqr) %>% 
  pivot_wider(
    names_from = config, 
    values_from = c(ci_50, iqr), 
    names_sep = "_"
  )

head(wide_metric_summary)

```
```{r}

wide_metric_summary <- wide_metric_summary %>% 
  mutate(
    delta_median_conc_vs_em = ci_50_concentration - ci_50_emission,
    delta_median_concem_vs_em = `ci_50_concentration-emission` - ci_50_emission,
    delta_median_conc_vs_concem = ci_50_concentration - `ci_50_concentration-emission`,

    spread_ratio_conc_vs_em = iqr_concentration / iqr_emission,
    spread_ratio_concem_vs_em = `iqr_concentration-emission` / iqr_emission,
    spread_ratio_conc_vs_concem = iqr_concentration / `iqr_concentration-emission`
  )

head(wide_metric_summary)
```

```{r}
# Reshape to long format for plotting delta medians
delta_long <- wide_metric_summary %>%
  select(scenario, term,
         delta_median_conc_vs_em,
         delta_median_concem_vs_em,
         delta_median_conc_vs_concem) %>%
  pivot_longer(
    cols = starts_with("delta_median"),
    names_to = "comparison",
    values_to = "delta_median"
  ) %>%
  mutate(comparison = recode(comparison,
    delta_median_conc_vs_em = "Concentration - Emission",
    delta_median_concem_vs_em = "Conc-Emission - Emission",
    delta_median_conc_vs_concem = "Concentration - Conc-Emission"
  ))

# Reshape to long format for plotting spread ratios
spread_long <- wide_metric_summary%>%
  select(scenario, term,
         spread_ratio_conc_vs_em,
         spread_ratio_concem_vs_em,
         spread_ratio_conc_vs_concem) %>%
  pivot_longer(
    cols = starts_with("spread_ratio"),
    names_to = "comparison",
    values_to = "spread_ratio"
  ) %>%
  mutate(comparison = recode(comparison,
    spread_ratio_conc_vs_em = "Concentration / Emission",
    spread_ratio_concem_vs_em = "Conc-Emission / Emission",
    spread_ratio_conc_vs_concem = "Concentration / Conc-Emission"
  ))
```

```{r}
# Reorder the term variable
delta_long$term <- factor(delta_long$term, levels = c("near-term", "mid-term", "long-term"))
spread_long$term <- factor(spread_long$term, levels = c("near-term", "mid-term", "long-term"))

# Plot delta medians
p1 <- ggplot(delta_long, aes(x = term, y = delta_median, group = comparison,
                             color = comparison)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Δ Median GMST Between Configurations",
    y = "Δ Median (°C)",
    x = "Time Period",
    color = "Comparison"
  ) +
  facet_wrap(~scenario) +
  theme_minimal()

# Plot spread ratios
p2 <- ggplot(spread_long, aes(x = term, y = spread_ratio, group = comparison,
                              color = comparison)) +
  geom_line() +
  geom_point() +
  labs(
    title = "IQR Ratio Between Configurations",
    y = "IQR Ratio",
    x = "Time Period",
    color = "Comparison"
  ) +
  facet_wrap(~scenario) +
  theme_minimal()

# Display plots
print(p1)
print(p2)
```

The above data as a heat map:
```{r}

# Use your previously computed wide_df with delta_median_* columns
heatmap_data <- wide_metric_summary %>%
  select(scenario, term,
         delta_median_conc_vs_em,
         delta_median_concem_vs_em,
         delta_median_conc_vs_concem) %>%
  pivot_longer(cols = starts_with("delta_median"),
               names_to = "comparison", values_to = "delta_median") %>%
  mutate(comparison = recode(comparison,
    delta_median_conc_vs_em = "Concentration - Emission",
    delta_median_concem_vs_em = "Conc-Emission - Emission",
    delta_median_conc_vs_concem = "Concentration - Conc-Emission"
  ))

# Optional: reorder term levels for clean X-axis
heatmap_data$term <- factor(heatmap_data$term, levels = c("near-term", "mid-term", "long-term"))

# Plot
ggplot(heatmap_data, aes(x = term, y = comparison, fill = delta_median)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(delta_median, 2)), size = 5, color = "darkgrey") +
  facet_wrap(~scenario) +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "YlGnBu"),
    name = "Delta Median (°C)"
  ) +
  labs(
    title = "Delta Median GMST Between Configurations",
    x = "Time Period",
    y = "Configuration Comparison",
    fill = "Delta Median (C)"
  ) +
  theme_light()

ggsave("median_heat_map.png", width = 11)
```

Uncertainty ratio figure:
```{r}
# Prepare heatmap data for IQR ratios
iqr_heatmap_data <- wide_metric_summary %>%
  select(scenario, term,
         spread_ratio_conc_vs_em,
         spread_ratio_concem_vs_em,
         spread_ratio_conc_vs_concem) %>%
  pivot_longer(
    cols = starts_with("spread_ratio"),
    names_to = "comparison",
    values_to = "spread_ratio"
  ) %>%
  mutate(comparison = recode(comparison,
    spread_ratio_conc_vs_em = "Concentration / Emission",
    spread_ratio_concem_vs_em = "Conc-Emission / Emission",
    spread_ratio_conc_vs_concem = "Concentration / Conc-Emission"
  ))

# Reorder term for X-axis
iqr_heatmap_data$term <- factor(iqr_heatmap_data$term, levels = c("near-term", "mid-term", "long-term"))

# Create IQR ratio heatmap (styled exactly like Delta Median)
ggplot(iqr_heatmap_data, aes(x = term, y = comparison, fill = spread_ratio)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(spread_ratio, 2)), size = 5, color = "darkgrey") +
  facet_wrap(~scenario) +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "YlGnBu"),
    name = "Very Likely Range Ratio"
  ) +
  labs(
    title = "VLR Ratio Between Configurations",
    x = "Time Period",
    y = "Configuration Comparison",
    fill = "VLR Ratio"
  ) +
  theme_light()

ggsave("uncertainty_heat_map.png", width = 11, height = 7)
```

