---
title: "Concentration Driven Hector + Matilda"
author: "Joe Brown"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

The goal for this file is to begin by running Hector + Matilda in concentration-driven mode. This means that we will be constraining history to CO2 concentrations and then using perturbed parameters to make future projections. 

This will be completed in two parts representing two experiments:

Experiment 1 - running Hector + Matilda prescribing CO2 concentrations for the historic and future time periods. 

Experiment 2 - running Hector + Matilda prescribing CO2 concentrations for the historic period and then allow emissions to take over after 2024.

# Set-up

## Load libraries

```{r, message=FALSE}

source("set-up/libraries.R")

```

## Load ini files

```{r}

source("set-up/load_ini.R")

```

## Load perturbed parameter data frame

This step loads a pre-generated perturbed parameter ensemble (PPE) that will be used across all Hector + Matilda experiments. The PPE was generated once using Hector defaults and saved to ensure consistent sampling across model configurations. Parameters are automatically "chunked" for parallel processing.

```{r}

source("set-up/load_params.R")

```

# Experiment 1 Running the model: Constrained to C02 in history & future

The goal here is to run Hector + Matilda in "concentration-driven" mode by constraining historical and future CO2 concentrations.

## Load CO2 data for constraint

```{r}

constraint_data_exp1 <- read.csv("data/processed/CO2_constrain_full.csv")

```

## Run model 

```{r}

source("set-up/utils.R")

```

Run Hector over entire PPE for each SSP scenario in the scenario list

```{r}
start_time <- Sys.time()
matilda_result_exp1 <- matilda_conc_driven(param_chunks, ini_list, constraint_data_exp1)
end_time <- Sys.time()

elapsed_time <- end_time - start_time
print(elapsed_time)

```

Correct the run_numbers for each of element in the result list

```{r}

matilda_result <- fix_run_numbers(matilda_result)

```

# Save model runs 

```{r}
# define directory path
result_dir <- here("output", "model_results")

# create directory if one does not exist
if (!dir.exists(result_dir)) {
  dir.create(result_dir, recursive = T)
}

# identify the file path
file_name <- "exp1_matilda_results_concentration_full.RDS"
file_path <- file.path(result_dir, file_name)

# save 
saveRDS(matilda_result, file_path)
```

# Experiment 2 Running the model: Constrained to C02 in history with emissions driven future

The goal here is to run Hector + Matilda in "concentration-driven" mode by constraining historical CO2 concentrations.

## Load CO2 data for constraint

```{r}

constraint_data_exp1 <- read.csv("data/processed/CO2_constrain_full.csv")

```

## Run model 

```{r}

source("set-up/utils.R")

```

Run Hector over entire PPE for each SSP scenario in the scenario list

```{r}
start_time <- Sys.time()
matilda_result_exp1 <- matilda_conc_driven(param_chunks, ini_list, constraint_data_exp1)
end_time <- Sys.time()

elapsed_time <- end_time - start_time
print(elapsed_time)

```

Correct the run_numbers for each of element in the result list

```{r}

matilda_result <- fix_run_numbers(matilda_result)

```

# Save model runs 

```{r}
# define directory path
result_dir <- here("output", "model_results")

# create directory if one does not exist
if (!dir.exists(result_dir)) {
  dir.create(result_dir, recursive = T)
}

# identify the file path
file_name <- "exp1_matilda_results_concentration_full.RDS"
file_path <- file.path(result_dir, file_name)

# save 
saveRDS(matilda_result, file_path)
```



# Quick visualization

```{r}
matilda_conc_driven <- do.call(rbind, matilda_result)
rownames(matilda_conc_driven) <- NULL

```

```{r}
ggplot(data = subset(matilda_conc_driven, variable %in% c("CO2_concentration", "global_tas"))) +
  geom_line(aes(x = year, y = value, group = run_number)) +
  facet_grid(variable ~ scenario, scales = "free_y") +
  theme_light()

```

