---
title: "Emissions Driven Hector + Matilda"
author: "Joe Brown"
date: "2025-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

The goal for this file is to begin by running Hector + Matilda in emissions-driven mode. This means that we will allow Hector to internally simulate historical CO2 concentrations from emissions inputs and then use perturbed parameters to make future projections.

# Set-up

## Load libraries

```{r, message=FALSE}

source("set-up/libraries.R")

```

## Load ini files

```{r}

source("set-up/load_ini.R")

```

## load perturbed parameter data frame

This step loads a pre-generated perturbed parameter ensemble (PPE) that will be used across all Hector + Matilda experiments. The PPE was generated once using Hector defaults and saved to ensure consistent sampling across model configurations. Parameters are automatically "chunked" for parallel processing.

```{r}

source("set-up/load_params.R")

```

# Running the model: Constrained to C02


## Run model 

```{r}

source("set-up/utils.R")

```

Run Hector over entire PPE for each SSP scenario in the scenario list

```{r}
start_time <- Sys.time()
matilda_result <- matilda_emission_driven(param_chunks, ini_list)
end_time <- Sys.time()

elapsed_time <- end_time - start_time
print(elapsed_time)

```

Correct the run_numbers for each of element in the result list

```{r}

matilda_result <- fix_run_numbers(matilda_result)

```

# Save model runs 

```{r}
# define directory path
result_dir <- here("output", "model_results")

# create directory if one does not exist
if (!dir.exists(result_dir)) {
  dir.create(result_dir, recursive = T)
}

# identify the file path
file_name <- "matilda_results_emission.RDS"
file_path <- file.path(result_dir, file_name)

# save
saveRDS(matilda_result, file_path)
```
