---
title: 'Testing Hypothesis #2'
author: "Joe Brown"
date: "2025-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

### Load libraries and utils 

```{r, message=FALSE}

source("set-up/libraries.R")
source("set-up/utils.R")

```

# Data 

Need Matilda warming projection results from each of the three configurations we are testing.

Load experiment GSAT metric results:
```{r}
lc_gsat_values_exp1_df <- read.csv("output/warming_results/exp1_gsat_metric.csv")
lc_gsat_values_exp1_df$config <- "CD"

lc_gsat_values_exp2_df <- read.csv("output/warming_results/exp2_gsat_metric.csv")
lc_gsat_values_exp2_df$config <- "CE"

lc_gsat_values_exp3_df <- read.csv("output/warming_results/exp3_gsat_metric.csv")
lc_gsat_values_exp3_df$config <- "ED"
```

Combine to a single data frame:
```{r}

lc_gsat_all <- rbind(lc_gsat_values_exp1_df, lc_gsat_values_exp2_df, lc_gsat_values_exp3_df)

```

Here we are preparing the data we need to conduct the KS test which will tell me whether the late century temperature distributions are different. 

```{r}
# split the lc_gsat_all into a list by config type
lc_config_list <- split(lc_gsat_all, lc_gsat_all$config)

# now split those config dfs again to split by scenario
lc_gsat_list <- lapply(lc_config_list, function(df) {
  split(df, df$scenario)
})

```

KS test:
```{r}
configs <- names(lc_gsat_list)
scenarios <- names(lc_gsat_list[[1]])  # assumes same scenarios in each config

# Define all config pairs you want to test
config_pairs <- combn(configs, 2, simplify = FALSE)

# Loop through each scenario and config pair
ks_results <- lapply(scenarios, function(scn) {
  lapply(config_pairs, function(pair) {
    
    config1 <- pair[1]
    config2 <- pair[2]
    
    x <- lc_gsat_list[[config1]][[scn]]$metric_result
    y <- lc_gsat_list[[config2]][[scn]]$metric_result
    
    ks <- ks.test(x, y)
    
    data.frame(
      scenario = scn,
      config1 = config1,
      config2 = config2,
      D_statistic = ks$statistic,
      p_value = ks$p.value
    )
    
  }) %>% bind_rows()
}) %>% bind_rows()
```

flag significance:
```{r}
ks_results <- ks_results %>%
  mutate(significant = p_value < 0.05)
```

write bootstrap function to get mean skewness or kurtosis:
```{r}
bootstrap_stat_moments <- function(values, stat_func, n_boot = 10000, n_sample = 1000) {
  replicate(n_boot, {
    resample <- sample(values, size = n_sample, replace = TRUE)
    stat_func(resample)
  })
}
```

Compute bootstrap fr distribution skew and kurtosis:
```{r}
set.seed(444)

# Bootstrap for skewness
bootstrap_skew_df <- lapply(names(lc_gsat_list), function(config) {
  
  config_data <- lc_gsat_list[[config]]
  
  lapply(names(config_data), function(scenario) {
    
    values <- config_data[[scenario]]$metric_result
    skew_vals <- bootstrap_stat_moments(values, skewness)

    data.frame(
      config = config,
      scenario = scenario,
      skewness = skew_vals
    )
    
  }) %>% bind_rows()
}) %>% bind_rows()

# Bootstrap for kurtosis
bootstrap_kurt_df <- lapply(names(lc_gsat_list), function(config) {
  
  config_data <- lc_gsat_list[[config]]
  
  lapply(names(config_data), function(scenario) {
    
    values <- config_data[[scenario]]$metric_result
    kurt_vals <- bootstrap_stat_moments(values, kurtosis)

    data.frame(
      config = config,
      scenario = scenario,
      kurtosis = kurt_vals
    )
    
  }) %>% bind_rows()
}) %>% bind_rows()


```

construct a summary tableusing bootstrap results
```{r}
# Skewness summary
skew_summary_df <- bootstrap_skew_df %>%
  group_by(config, scenario) %>%
  summarize(
    mean_skew = mean(skewness),
    skew_ci_low = quantile(skewness, 0.025),
    skew_ci_high = quantile(skewness, 0.975),
    .groups = "drop"
  )

# Kurtosis summary
kurt_summary_df <- bootstrap_kurt_df %>%
  group_by(config, scenario) %>%
  summarize(
    mean_kurt = mean(kurtosis),
    kurt_ci_low = quantile(kurtosis, 0.025),
    kurt_ci_high = quantile(kurtosis, 0.975),
    .groups = "drop"
  )
```

Plotting:
```{r}
skew_plot <- ggplot(data = skew_summary_df) +
  geom_bar(aes(x = config, y = mean_skew, fill = config), stat = "identity") +
  geom_errorbar(aes(x = config, ymin = skew_ci_low, ymax = skew_ci_high), width = 0.25) +
  facet_wrap(~scenario) +
  theme_light()

ggsave("output/figures/skew_plot.png", skew_plot, 
       height = 5, 
       width = 10, 
       units = "in", 
       device = "png")
```

```{r}

kurt_plot <- ggplot(data = kurt_summary_df) +
  geom_bar(aes(x = config, y = mean_kurt, fill = config), stat = "identity") +
  geom_errorbar(aes(x = config, ymin = kurt_ci_low, ymax = kurt_ci_high), width = 0.25) +
  facet_wrap(~scenario) +
  theme_light()

ggsave("output/figures/kurt_plot.png", kurt_plot, 
       height = 5, 
       width = 10, 
       units = "in", 
       device = "png")
```

